---
title: "Ecological Genomics: Planning Translocations"
subtitle: "Simulating genetic drift and building an interactive Shiny app"
author: "Daniel Croll"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: false
    code-tools: true
    embed-resources: true
---

## Learning Objectives

By the end of this tutorial you will be able to:

- Explain why genetic drift erodes heterozygosity in small, isolated populations
- Simulate allele-frequency trajectories across multiple populations in R using `learnPopGen`
- Compute and visualise the expected and observed loss in heterozygosity over time
- Build an interactive Shiny app that lets a conservation manager explore translocation scenarios

---

## Submission of your work

You can work in groups of two or more people. 

Submit a succinct report including responses to the questions in the tutorial. You can optionally include screenshots of particular plots or app outputs to illustrate your answers. The focus should be on simple explanations and exploration of your findings.

Each student should submit on Moodle their own copy of the responses to the questions in the tutorial, but you can collaborate on the exploration of the code and discuss responses. 

---

## Background: Conservation Genomics and Translocations

Small, isolated populations lose genetic diversity over time through **genetic drift**. This matters in conservation because lower heterozygosity is potentially associated with reduced fitness, impaired immune response, and a decreased ability to adapt to environmental change.

**Translocation** is the deliberate movement of individuals between populations to:

1. Rescue a genetically depauperate population (*genetic rescue*)
2. Reintroduce a species to a recovered habitat
3. Buffer a population against inbreeding depression

Before carrying out a translocation, managers need to anticipate *how much* genetic diversity a target population is likely to have lost given its history. That is exactly what this simulator will help you estimate.

The expected heterozygosity at generation $t$ under pure drift in a closed population of effective size $N_e$ is:

$$H_t = H_0 \left(1 - \frac{1}{2N_e}\right)^t$$

where $H_0 = 2p_0(1-p_0)$ is the initial expected heterozygosity.

---

## Part 1: Setting Up Your Environment

Before you start, make sure the required packages are installed. **NB: this is not necessary on the server.**

If needed, run the following chunk **once** (you can set `eval: false` after the first run).

```{r}
#| label: install
#| eval: false
install.packages("learnPopGen")
install.packages("ggplot2")
install.packages("tidyr")
install.packages("dplyr")
install.packages("shiny")
```

Load the packages:

```{r}
#| label: setup
#| message: false
#| warning: false
library(learnPopGen)
library(ggplot2)
library(tidyr)
library(dplyr)
```

---

## Part 2: Simulating Genetic Drift in 5 Isolated Populations

### Running the simulation

The `learnPopGen` package provides the function `genetic.drift()`, which simulates random genetic drift at a single bi-allelic locus across replicate populations.

Key arguments:

| Argument | Meaning |
|----------|---------|
| `p0`     | Initial frequency of allele A (0–1) |
| `Ne`     | Effective population size |
| `nrep`   | Number of replicate populations |
| `time`   | Number of generations to simulate |
| `show`   | What to plot: `"p"` for allele frequency |

Run the simulation with **5 populations** (`nrep = 5`) of equal effective size:

```{r}
#| label: simulate-drift
set.seed(123)          # For reproducibility

p0   <- 0.5            # Starting allele frequency
Ne   <- 100            # Effective population size (same for all 5 pops)
ngen <- 200            # Number of generations

# genetic.drift() returns a 'genetic.drift' S3 object.
# unclass() strips the custom class so it behaves as a plain numeric matrix,
# which is required by as.data.frame() and other standard operations.
# pause = 0 disables the animation delay (speeds up rendering).
freq_matrix <- unclass(genetic.drift(
  p0    = p0,
  Ne    = Ne,
  nrep  = 5,
  time  = ngen,
  show  = "p",
  pause = 0
))

cat("Dimensions of the frequency matrix:", dim(freq_matrix), "\n")
cat("(rows = generations, columns = populations)\n\n")
cat("Allele frequencies at generation 1 (starting values):\n")
print(freq_matrix[1, ])
```

> **Q1** - Inspect the columns of `freq_matrix`. What does each column represent?

### Computing heterozygosity

Expected heterozygosity at a single locus with allele frequency $p$ is $H = 2p(1-p)$. Apply this to every cell of the frequency matrix:

```{r}
#| label: compute-H
# Compute heterozygosity for each population at each generation
H_matrix <- 2 * freq_matrix * (1 - freq_matrix)

# Compute the mean heterozygosity across all 5 populations at each generation
H_mean <- rowMeans(H_matrix)

cat("H at generation 1 (all pops start at p0 =", p0, "):", round(H_matrix[1, 1], 4), "\n")
cat("Mean H at final generation (", ngen, "):", round(H_mean[ngen], 4), "\n")
```

Add the **predicted heterozygosity** from the allele frequencies:

```{r}
#| label: theory
generations <- 1:ngen
H0          <- 2 * p0 * (1 - p0)          # Initial heterozygosity
H_expected  <- H0 * (1 - 1 / (2 * Ne))^generations
```

---

## Part 3: Visualising Heterozygosity Loss

Build a tidy data frame so we can plot everything cleanly with `ggplot2`:

```{r}
#| label: tidy-data
# unclass() ensures H_matrix is treated as a plain numeric matrix by
# as.data.frame() (the genetic.drift S3 class has a custom method that
# would otherwise melt the matrix into a single-column data frame)
H_df <- as.data.frame(unclass(H_matrix))
colnames(H_df) <- paste0("Pop", 1:5)
H_df$generation <- generations

H_long <- pivot_longer(
  H_df,
  cols      = starts_with("Pop"),
  names_to  = "population",
  values_to = "heterozygosity"
)

# Add the theoretical expectation to every row so each population panel
# can overlay its own expected trajectory
H_long <- H_long %>%
  left_join(
    data.frame(generation = generations, H_expected = H_expected),
    by = "generation"
  )

# Summary data frame for the overview plot (observed mean + theory)
summary_df <- data.frame(
  generation     = generations,
  H_mean         = H_mean,
  H_expected     = H_expected
)
```

First, an **overview** showing all population trajectories together with the observed mean and the prediction:

```{r}
#| label: plot-H-overview
#| fig-width: 8
#| fig-height: 5
ggplot() +
  # Individual population trajectories (thin, semi-transparent)
  geom_line(
    data  = H_long,
    aes(x = generation, y = heterozygosity, colour = population),
    alpha = 0.5, linewidth = 0.6
  ) +
  # Mean across populations
  geom_line(
    data   = summary_df,
    aes(x  = generation, y = H_mean),
    colour = "black", linewidth = 1.2, linetype = "solid"
  ) +
  # Theoretical expectation
  geom_line(
    data   = summary_df,
    aes(x  = generation, y = H_expected),
    colour = "red", linewidth = 1, linetype = "dashed"
  ) +
  annotate("text", x = ngen * 0.68, y = H_expected[round(ngen * 0.68)] * 1.07,
           label = "Expected H", colour = "red", hjust = 0, size = 3.5) +
  annotate("text", x = ngen * 0.68, y = H_mean[round(ngen * 0.68)] * 0.88,
           label = "Observed mean H", colour = "black", hjust = 0, size = 3.5) +
  labs(
    title  = paste0("Overview: heterozygosity loss in 5 isolated populations (Ne = ", Ne, ")"),
    x      = "Generation",
    y      = "Heterozygosity (H)",
    colour = "Population"
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "right")
```

Now **facet by population** so we can compare the observed and expected trajectories for each population individually. The red dashed line is the same theoretical prediction in every panel — deviations from it reflect stochastic drift unique to that population:

```{r}
#| label: plot-H-per-pop
#| fig-width: 9
#| fig-height: 6
ggplot(H_long, aes(x = generation)) +
  # Observed heterozygosity for this population
  geom_line(
    aes(y = heterozygosity, colour = population),
    linewidth = 0.7
  ) +
  # Theoretical expectation (same for every panel)
  geom_line(
    aes(y = H_expected),
    colour = "red", linewidth = 0.8, linetype = "dashed"
  ) +
  facet_wrap(~ population, ncol = 3) +
  labs(
    title   = paste0("Observed (solid) vs. expected (red dashed) heterozygosity per population (Ne = ", Ne, ")"),
    x       = "Generation",
    y       = "Heterozygosity (H)",
    colour  = "Population"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "grey92", colour = NA))
```

> **Q2** – How does the observed heterozygosity loss compare to the expectation? Is one higher than the other? Compare populations.

> **Q3** – Repeat the simulation with `Ne = 20` and `Ne = 500`. You can adjust this in the first code block of **Running the simulation**, then repeat the steps until the visualization. How does the rate of heterozygosity loss change?

---

## Part 4: Simulating a Translocation Event

After a period of genetic drift, each population has a different allele frequency — some may have drifted close to fixation (p ≈ 0 or p ≈ 1), drastically reducing their heterozygosity. A **translocation** introduces migrants from a source population (e.g., a reservoir population or ex-situ captive group), shifting allele frequencies by restoring some heterozygosity.

### Building a custom popgen simulator

`genetic.drift()` from `learnPopGen` simulates the entire run in one call and returns the complete trajectory, so there is no way to pause it mid-run and inject migrants. We therefore write our own equivalent function using **binomial sampling** — the same biological process `genetic.drift()` implements internally:

At each generation, the number of copies of allele A out of $2N_e$ drawn is:

$$k_t \sim \text{Binomial}(2N_e,\; p_{t-1})$$

$$p_t = \frac{k_t}{2N_e}$$

```{r}
#| label: simulate-drift-fn
# Custom genetic drift simulator
# Returns a matrix of allele frequencies: rows = generations, columns = populations
simulate_drift <- function(p_start, Ne, nrep, ngen) {
  freq        <- matrix(NA, nrow = ngen, ncol = nrep)
  freq[1, ]   <- p_start
  for (t in 2:ngen) {
    allele_counts <- rbinom(nrep, size = 2 * Ne, prob = freq[t - 1, ])
    freq[t, ]     <- allele_counts / (2 * Ne)
  }
  freq
}
```

Verify that it gives the same qualitative behaviour as `genetic.drift()`:

```{r}
#| label: verify-custom-fn
set.seed(123)
freq_matrix <- simulate_drift(p_start = 0.5, Ne = 100, nrep = 5, ngen = 200)
cat("Dimensions:", dim(freq_matrix), "\n")
cat("Starting frequencies (gen 1):", freq_matrix[1, ], "\n")
```

### The allele-frequency mixing formula

When $n_m$ migrants from a **source** with allele frequency $p_{\text{source}}$ enter a **target** population of size $N_e$ at generation $t^*$, the new allele frequency is a weighted average:

$$p_{t^*}^{\,\text{new}} = \frac{N_e \cdot p_{t^*} + n_m \cdot p_{\text{source}}}{N_e + n_m}$$

This immediately pushes $p$ closer to $p_{\text{source}}$. Because $H = 2p(1-p)$ is maximised at $p = 0.5$, any movement of a near-fixed population (p ≈ 0 or p ≈ 1) toward 0.5 increases heterozygosity.

### Two-phase simulation

We split the simulation at the translocation generation $t^*$:

- **Phase 1** — simulate drift for $t^*$ generations (same as before)
- **Apply mixing** — update allele frequencies using the formula above
- **Phase 2** — continue drift from the new frequencies for the remaining $t_{\text{gen}} - t^*$ generations

```{r}
#| label: translocation-sim
set.seed(123)

# ── Parameters ──────────────────────────────────────────────────────────────
p0         <- 0.5    # Initial allele frequency
Ne         <- 50    # Effective population size
nrep       <- 5      # Number of populations
ngen       <- 200    # Total generations
t_trans    <- 80     # Generation at which translocation occurs
n_migrants <- 10     # Number of migrants introduced
p_source   <- 0.5    # Allele frequency in the source population

# ── Phase 1: drift to t_trans ────────────────────────────────────────────────
freq_pre  <- simulate_drift(p0, Ne, nrep, t_trans)

# Allele frequencies and heterozygosity immediately BEFORE translocation
p_before  <- freq_pre[t_trans, ]
H_before  <- 2 * p_before * (1 - p_before)

cat("Allele frequencies at generation", t_trans, "(before translocation):\n")
print(round(p_before, 3))
cat("\nHeterozygosity per population BEFORE translocation:\n")
print(round(H_before, 3))

# ── Apply allele-frequency mixing ────────────────────────────────────────────
p_after <- (Ne * p_before + n_migrants * p_source) / (Ne + n_migrants)
H_after <- 2 * p_after * (1 - p_after)

cat("\nAllele frequencies directly AFTER translocation:\n")
print(round(p_after, 3))
cat("\nHeterozygosity per population AFTER translocation:\n")
print(round(H_after, 3))
cat("\nAbsolute mean H gain at translocation:",
    round(mean(H_after) - mean(H_before), 4), "\n")

# ── Phase 2: continue drift from post-translocation frequencies ──────────────
freq_post     <- simulate_drift(p_after, Ne, nrep, ngen - t_trans + 1)
# Combine: row t_trans of freq_post[1,] duplicates freq_pre[t_trans,] → drop it
freq_with_trans <- rbind(freq_pre, freq_post[-1, ])

# ── Baseline: same simulation WITHOUT translocation ──────────────────────────
set.seed(123)
freq_no_trans <- simulate_drift(p0, Ne, nrep, ngen)
```

### Comparing scenarios

```{r}
#| label: translocation-compare
#| fig-width: 13
#| fig-height: 5
generations <- 1:ngen
H0          <- 2 * p0 * (1 - p0)
H_expected  <- H0 * (1 - 1 / (2 * Ne))^generations

# Mean H trajectories for both scenarios
H_mean_with <- rowMeans(2 * freq_with_trans * (1 - freq_with_trans))
H_mean_no   <- rowMeans(2 * freq_no_trans   * (1 - freq_no_trans))

# Per-population H in long format (with translocation only)
H_mat_with <- 2 * freq_with_trans * (1 - freq_with_trans)
H_df_with  <- as.data.frame(H_mat_with)
colnames(H_df_with) <- paste0("Pop", 1:nrep)
H_df_with$generation <- generations
H_long_with <- pivot_longer(H_df_with, starts_with("Pop"),
                             names_to = "population", values_to = "H_observed")
H_long_with$scenario <- "With translocation"

H_mat_no <- 2 * freq_no_trans * (1 - freq_no_trans)
H_df_no  <- as.data.frame(H_mat_no)
colnames(H_df_no) <- paste0("Pop", 1:nrep)
H_df_no$generation <- generations
H_long_no <- pivot_longer(H_df_no, starts_with("Pop"),
                           names_to = "population", values_to = "H_observed")
H_long_no$scenario <- "No translocation"

H_long_both <- bind_rows(H_long_with, H_long_no)

# Mean lines + expected
summary_comparison <- data.frame(
  generation = rep(generations, 3),
  H          = c(H_mean_with, H_mean_no, H_expected),
  line       = rep(c("Observed mean — with translocation",
                     "Observed mean — no translocation",
                     "Expected (no translocation)"), each = ngen)
)

ggplot() +
  # Per-population lines, coloured by scenario
  geom_line(
    data  = H_long_both,
    aes(x = generation, y = H_observed,
        colour = scenario, group = interaction(population, scenario)),
    alpha = 0.35, linewidth = 0.55
  ) +
  # Mean lines
  geom_line(
    data = summary_comparison,
    aes(x = generation, y = H, linetype = line, colour = line),
    linewidth = 1
  ) +
  # Translocation marker
  geom_vline(xintercept = t_trans, linetype = "dotted", colour = "grey30") +
  annotate("text", x = t_trans + 3, y = H0 * 0.97,
           label = paste0("Translocation\n(gen ", t_trans, ", ", n_migrants, " migrants)"),
           hjust = 0, size = 3.2, colour = "grey30") +
  scale_colour_manual(
    values = c(
      "With translocation"                   = "#2196F3",
      "No translocation"                     = "#FF7043",
      "Observed mean — with translocation"   = "#0D47A1",
      "Observed mean — no translocation"     = "#BF360C",
      "Expected (no translocation)"          = "darkred"
    ),
    guide = guide_legend(
      override.aes = list(
        # alphabetical order: Expected, No translocation, Obs mean no, Obs mean with, With translocation
        linetype = c("dashed", "solid", "solid", "solid", "solid"),
        linewidth = c(1, 0.55, 1, 1, 0.55),
        alpha    = c(1, 0.35, 1, 1, 0.35)
      )
    )
  ) +
  scale_linetype_manual(values = c(
    "With translocation"                  = "solid",
    "No translocation"                    = "solid",
    "Observed mean — with translocation"  = "solid",
    "Observed mean — no translocation"    = "solid",
    "Expected (no translocation)"         = "dashed"
  )) +
  guides(linetype = "none") +
  labs(
    title    = paste0("Effect of translocation on heterozygosity (Ne = ", Ne,
                      ", ", n_migrants, " migrants at gen ", t_trans, ")"),
    x        = "Generation",
    y        = "Heterozygosity (H)",
    colour   = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "right",
        legend.key.width = unit(1.5, "cm"))
```

> **Q4** – Using the set values, did the translocation increase heterozygosity?

---

## Part 5: Building Your Own Shiny App

A static script is useful, but a conservation manager without R knowledge cannot explore scenarios interactively. A **Shiny app** lets you turn your simulation into a browser-based tool with sliders and live plots. Here, we will build such an app.

### How Shiny works

Every Shiny app is a single R script (`app.R`) with three elements:

```r
library(shiny)

ui <- fluidPage(          # 1. Layout + inputs + output placeholders
  ...
)

server <- function(input, output) {   # 2. Logic: reads inputs, renders outputs
  output$myPlot <- renderPlot({ ... })
}

shinyApp(ui = ui, server = server)    # 3. Launches the app
```
d
Key concepts:

| Concept | What it does |
|---|---|
| `sliderInput("id", ...)` | Creates a slider; its value is read as `input$id` in the server |
| `checkboxInput("id", ...)` | Creates a checkbox; value is `TRUE`/`FALSE` |
| `plotOutput("id")` | Reserves a panel for a plot |
| `renderPlot({ })` | Runs whenever any `input$` it uses changes; returns a ggplot |
| `reactive({ })` | Caches a computed result and re-runs only when its inputs change |

> **RStudio Server note:** Always view your app in the **Viewer pane** (bottom-right). Do *not* click "Open in Browser" — that opens `localhost:PORT` on your local machine, which cannot reach the remote server.

### Step-by-step: create `app.R`

In RStudio, go to **File > New File > R Script**, immediately save it as `app.R` in your home folder, then build it up following the steps below.

Alternatively, you can download the complete `app.R` file from the GitHub repository:

```{r}

```
download.file(
  "https://raw.githubusercontent.com/crolllab/teaching/master/Ecological_Genomics/app.R",
  destfile = "app.R",
  method   = "wget"
)
```

**Step 1 — Load packages and paste `simulate_drift()`**

Your app needs the same custom simulator from Part 4. Copy it at the top of `app.R`:

```r
library(shiny)
library(ggplot2)
library(tidyr)
library(dplyr)

simulate_drift <- function(p_start, Ne, nrep, ngen) {
  freq      <- matrix(NA, nrow = ngen, ncol = nrep)
  freq[1, ] <- p_start
  for (t in 2:ngen) {
    freq[t, ] <- rbinom(nrep, size = 2 * Ne, prob = freq[t - 1, ]) / (2 * Ne)
  }
  freq
}
```

**Step 2 — Build the UI**

Create a sidebar layout with sliders for the population parameters and the translocation event. Replace each `# TODO` with the appropriate `sliderInput()` or `numericInput()` call:

```r
ui <- fluidPage(
  titlePanel("Translocation Planner: Heterozygosity Simulator"),
  sidebarLayout(
    sidebarPanel(
      h4("Population parameters"),
      sliderInput("Ne",     "Effective population size (Ne):",
                  min = 10, max = 500, value = 100, step = 10),
      sliderInput("n_pops", "Number of populations:",
                  min = 2,  max = 10,  value = 5,   step = 1),
      sliderInput("ngen",   "Total generations:",
                  min = 50, max = 500, value = 200, step = 10),
      numericInput("p0",    "Initial allele frequency (p0):",
                   min = 0.05, max = 0.95, value = 0.5, step = 0.05),
      hr(),
      h4("Translocation event"),
      checkboxInput("do_trans", "Enable translocation", value = TRUE),
      sliderInput("t_trans",    "Generation of translocation:",
                  min = 10, max = 490, value = 80, step = 5),
      sliderInput("n_migrants", "Number of migrants:",
                  min = 1,  max = 100, value = 10, step = 1),
      numericInput("p_source",  "Source allele frequency:",
                   min = 0.05, max = 0.95, value = 0.5, step = 0.05)
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("Heterozygosity over time", plotOutput("hetPlot", height = "460px")),
        tabPanel("Summary table",            tableOutput("summaryTable"))
      )
    )
  )
)
```

**Step 3 — Write the server**

The server runs both drift scenarios and renders the plot:

```r
server <- function(input, output) {

  sim_data <- reactive({
    set.seed(42)
    Ne      <- input$Ne
    nrep    <- input$n_pops
    ngen    <- input$ngen
    p0      <- input$p0
    t_trans <- min(input$t_trans, ngen - 1)   # guard: must be < ngen
    n_m     <- input$n_migrants
    p_src   <- input$p_source

    # Baseline: pure drift (no translocation)
    freq_no <- simulate_drift(p0, Ne, nrep, ngen)

    # With translocation (two-phase)
    freq_pre  <- simulate_drift(p0, Ne, nrep, t_trans)
    p_after   <- (Ne * freq_pre[t_trans, ] + n_m * p_src) / (Ne + n_m)
    freq_post <- simulate_drift(p_after, Ne, nrep, ngen - t_trans + 1)
    freq_with <- rbind(freq_pre, freq_post[-1, ])

    list(freq_no = freq_no, freq_with = freq_with, t_trans = t_trans)
  })

  output$hetPlot <- renderPlot({
    dat         <- sim_data()
    generations <- 1:input$ngen
    H0          <- 2 * input$p0 * (1 - input$p0)
    H_th        <- H0 * (1 - 1 / (2 * input$Ne))^generations

    make_long <- function(freq_mat, label) {
      H_mat <- 2 * freq_mat * (1 - freq_mat)
      df    <- as.data.frame(H_mat)
      colnames(df) <- paste0("Pop", seq_len(input$n_pops))
      df$generation <- generations
      long <- pivot_longer(df, starts_with("Pop"),
                           names_to = "population", values_to = "H_observed")
      long$scenario <- label
      long
    }

    H_long_both <- bind_rows(
      make_long(dat$freq_no,   "No translocation"),
      make_long(dat$freq_with, "With translocation")
    )

    H_mean_no   <- rowMeans(2 * dat$freq_no   * (1 - dat$freq_no))
    H_mean_with <- rowMeans(2 * dat$freq_with * (1 - dat$freq_with))

    means_df <- data.frame(
      generation = rep(generations, 3),
      H    = c(H_mean_no, H_mean_with, H_th),
      line = rep(c("Observed mean — no translocation",
                   "Observed mean — with translocation",
                   "Expected (no translocation)"), each = input$ngen)
    )

    p <- ggplot() +
      geom_line(
        data  = H_long_both,
        aes(x = generation, y = H_observed,
            colour = scenario, group = interaction(population, scenario)),
        alpha = 0.35, linewidth = 0.55
      ) +
      geom_line(
        data = means_df,
        aes(x = generation, y = H, linetype = line, colour = line),
        linewidth = 1
      ) +
      scale_colour_manual(
        values = c(
          "With translocation"                   = "#2196F3",
          "No translocation"                     = "#FF7043",
          "Observed mean — with translocation"   = "#0D47A1",
          "Observed mean — no translocation"     = "#BF360C",
          "Expected (no translocation)"          = "darkred"
        ),
        guide = guide_legend(
          override.aes = list(
            linetype  = c("dashed", "solid", "solid", "solid", "solid"),
            linewidth = c(1, 0.55, 1, 1, 0.55),
            alpha     = c(1, 0.35, 1, 1, 0.35)
          )
        )
      ) +
      scale_linetype_manual(values = c(
        "With translocation"                   = "solid",
        "No translocation"                     = "solid",
        "Observed mean — with translocation"   = "solid",
        "Observed mean — no translocation"     = "solid",
        "Expected (no translocation)"          = "dashed"
      )) +
      guides(linetype = "none") +
      labs(
        title  = paste0("Heterozygosity over time (Ne = ", input$Ne,
                        ", ", input$n_migrants, " migrants at gen ", dat$t_trans, ")"),
        x      = "Generation",
        y      = "Heterozygosity (H)",
        colour = NULL
      ) +
      theme_classic(base_size = 13) +
      theme(legend.position = "right",
            legend.key.width = unit(1.5, "cm"))

    if (input$do_trans) {
      p <- p +
        geom_vline(xintercept = dat$t_trans, linetype = "dotted", colour = "grey30") +
        annotate("text", x = dat$t_trans + 2, y = H0 * 0.97,
                 label = paste0("Translocation\n(gen ", dat$t_trans,
                                ", ", input$n_migrants, " migrants)"),
                 hjust = 0, size = 3.2, colour = "grey30")
    }
    p
  })

  output$summaryTable <- renderTable({
    dat  <- sim_data()
    ngen <- input$ngen
    H0   <- 2 * input$p0 * (1 - input$p0)

    H_final_no   <- mean(2 * dat$freq_no[ngen, ]   * (1 - dat$freq_no[ngen, ]))
    H_final_with <- mean(2 * dat$freq_with[ngen, ] * (1 - dat$freq_with[ngen, ]))
    H_th_final   <- H0 * (1 - 1 / (2 * input$Ne))^ngen

    data.frame(
      Statistic = c(
        "Initial heterozygosity (H0)",
        "Expected H at final generation (theory)",
        "Observed mean H — no translocation",
        "Observed mean H — with translocation",
        "Absolute H rescued by translocation",
        "% of H0 retained — no translocation",
        "% of H0 retained — with translocation"
      ),
      Value = round(c(
        H0,
        H_th_final,
        H_final_no,
        H_final_with,
        H_final_with - H_final_no,
        100 * H_final_no   / H0,
        100 * H_final_with / H0
      ), 4)
    )
  })
}

shinyApp(ui = ui, server = server)
```

**Step 3 — Save file and run the app**

Make sure to save `app.R`, then click the green "Run App" button in RStudio. The app should launch in the Viewer pane. Adjust the sliders and checkboxes to explore different scenarios.

> **Q5** – Explore the parameter space (population size, number of migrants, allele frequencies in source and recipient populations, etc.). Get a feeling for the impact of the timing and magnitude of the translocation on the heterozygosity trajectories. Then, imagine a particular, highly endangered species and conservation scenario. Define some plausible parameter values for population size and number of populations. Explore what translocation strategy (timing, number of migrants, source allele frequency) would maximise the heterozygosity rescue for that species. Investigate your scenarios and discuss your findings how well a translocation program could mitigate the loss of heterozygosity in your chosen case.

